---
title: "Specific agreement on dichotomous outcomes in the situation of more than two raters"
author: "Iris Eekhout"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agreement on dichotomous outcomes in the situation of more than two raters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document describes the use of the Agree package for the data example that was used in the paper on specific agreement on dichotomous outcomes in the situation of more than two raters.  

```{r}
install.packages("devtools")
devtools::install_github("iriseekhout/Agree")
library(Agree)
```

## Description of data example
For the data example we used data from a study by Dikmans et al. (2015). This data is based on photographs’s of breasts of 50 women after breast reconstruction. The photographs are  independently scored by 5 surgeons, the patient, and three mothers. They each rated the quality of the reconstruction on a 5 point ordinal scale with the verbal anchors on the left side ‘very dissatisfied’ on the left end and on the right end ‘very satisfied’ on the right end. They specifically rated the volume, shape, symmetry, scars and nipple. In this paper we use the data of 4 surgeons because one surgeon had some missing values and we look at the rates for symmetry. The satisfaction scores were dichotomized into satistified (scores 4 and 5) and not satistified (scores 1,2, and 3).

```{r}
load("data/breast.rda")
variable <- "symmetry"
raters <- c("PCH1", "PCH2", "PCH3", "PCH4")
ratersvars <- paste(raters, variable, sep="_")
data1 <- data.frame(breast[ratersvars])
 for (r in 1:length(ratersvars)){
    data1[ratersvars[r]] <- ifelse(data1[ratersvars[r]]=="very satisfied"|data1[ratersvars[r]]=="satisfied","satisfied","not satisfied")
 }
data1 <- data.frame(apply(data1[ratersvars],2,as.factor))
head(data1)
```

### Agreement table
First the agreement tables are summed for all rater combinations into one agreement table. Then the off diagonal cells are averaged to obtain symmetry agreement tables. Note that data1 contains a column per rater for the variale of interest.

```{r}
sumtable(data1,offdiag = FALSE)
sumtable(data1,offdiag = TRUE)
```

### Agreement
From the agreement table we can calculate the agreement. And we can calculate the confidence interval around this agreement.
```{r}
agreement(data1)
CIagreement(data1,m=4,n=nrow(data1))

```


### Specific agreement
The specific agreement for dichotomous data can be evaluated for the satisfied scores and for the not satisfied scores. 
```{r}
specific.agreement(data1, cat1="satisfied")
CIagreement(data1, cat1="satisfied")


specific.agreement(data1, cat1="not satisfied")
CIagreement(data1, cat1="not satisfied")

```

### Validation of Confidence interval
Simulation to compare the Fleis correction

Follows....
