---
title: "Specific agreement on polytomous outcomes in the situation of more than two raters"
author: "Iris Eekhout"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Agreement on polytomous outcomes in the situation of more than two raters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This document describes the use of the Agree package for two data examples dat are used in the paper on specific agreement on polytomous outcomes in the situation of more than two raters. The first data example is an example of ordinal ratings and the second example of nominal rating. 

```{r}
library(Agree)
library(pander)
```

## Ordinal data example
For the ordinal data example we use data from a study by Dikmans et al. (2015). This data is based on photographs’s of breasts of 50 women after breast reconstruction. The photographs are  independently scored by 5 surgeons, the patien, and three mothers. They each rated the quality of the reconstruction on a 5 point ordinal scale with the verbal anchors on the left side ‘very dissatisfied’ on the left end and on the right end ‘very satisfied’ on the right end. They specifically rated the volume, shape, symmetry, scars and nipple. In this paper we use the data of 4 surgeons because one surgeon had some missing values and we look at the rates for symmetry. Data set 1 is used for the example of ordinal categories.

```{r}
load("data/breast.rda")

variable <- "symmetry"
raters <- c("PCH1", "PCH2", "PCH3", "PCH4")
ratersvars <- paste(raters, variable, sep="_")
data1 <- data.frame(breast[ratersvars])
```

### Agreement table
First the agreement table are summed for all rater combinations into one agreement table. Then the off diagonal cells are averaged to obtain symmetry agreement tables. 

```{r}
pander(sumtable(data1,offdiag = FALSE))
pander(sumtable(data1,offdiag = TRUE))
```

### Agreement
From the agreement table we can calculate the agreement. And we can calculate the confidence interval around this agreement.
```{r}
agreement(data1)
CIagreement(data1,m=4,n=nrow(data1))

```


### Specific agreement
The specific agreement for polytomous data, can be defined in two ways: the agreement of for one category versus not that category (e.g. very satisfied versus all other categories) or the agreement for one category versus any other (e.g. very satistfied versus satisfied). The function below now returns the folowing output:

- A table (i.e. "betweentable") that is the summed agreement table with averaged discordant cells, with an extra column containing the row sums and an extra row containing the column proportions. 
- The proportion for each category, which is the column proportions from the between table, seperately returned. 
- The proportions for each cell are returned in the cel.proportions table.
- The specific agreement is in this function the agreement for one category versus all others.

```{r}
specific.agreement(data1, cat1="satisfied")
CIagreement(data1, cat1="satisfied")


specific.agreement(data1, cat1="satisfied", cat2="very satisfied")
CIagreement(data1, cat1="satisfied", cat2="very satisfied")


specific.agreement(data1, cat1="satisfied", cat2="neutral")
CIagreement(data1, cat1="satisfied", cat2="neutral")

```


###Conditional probability
We can calulate the probability of any other outcome conditional on an specific outcome. 

```{r}
conditional.agreement(data1)
```


### Weighted agreement
For ordinal data it might also be useful to look at the agreement when they may be one category off. So the agreement plus or minus one category. 

```{r}
weighted.agreement(data1)

weighted.agreement(data1, weight=0.5)

```



## nominal data example
For the nominal data example we use a dataset that was used in a paper by Fleis (1971), which originated from Sandifer et al. In this data patients are diagnosed in 5 categories: Depession, Personality Disorder, Schizophrenia, Neurois, and Other by 6 raters. 

```{r}

load("data/diagnoses.rda")
data2 <- data.frame(lapply(diagnoses,as.factor))
```


### Agreement table
First the agreement table are summed for all rater combinations into one agreement table. Then the off diagonal cells are averaged to obtain symmetry agreement tables. 

```{r}
pander(sumtable(data2,offdiag = FALSE))
pander(sumtable(data2,offdiag = TRUE))
```

### Agreement
From the agreement table we can calculate the agreement. And we can calculate the confidence interval around this agreement.
```{r}
agreement(data2)
CIagreement(data2)
```

### Specific agreement
The specific agreement for polytomous data, can be defined in two ways: the agreement of for one category versus not that category (e.g. Depression versus all other categories) or the agreement for one category versus any other (e.g. Depression versus Schizophrenia). 
```{r}
specific.agreement(data2, cat1="1. Depression")
CIagreement(data2, cat1="1. Depression")

specific.agreement(data2, cat1="2. Personality Disorder")
CIagreement(data2, cat1="2. Personality Disorder")

specific.agreement(data2, cat1="3. Schizophrenia")
CIagreement(data2, cat1="3. Schizophrenia")

specific.agreement(data2, cat1="4. Neurosis")
CIagreement(data2, cat1="4. Neurosis")

specific.agreement(data2, cat1="5. Other")
CIagreement(data2, cat1="5. Other")

```


### Conditional agreement

```{r}
conditional.agreement(data2)
```

